---
title: Dec time
description: >
  Introducing Declock, a timekeeping system that displays time in decimal days using math notation without the need for hours, minutes, or seconds.
image: /asset/daywide.svg
url: https://maptv.github.io/dec/time
decdate: 2026+222
draft: true
slug: time
aliases:
  - /time
  - /dt
engine: knitr
---

::: {.timenav}
{{< include /asset/_decnav.qmd >}}
:::

# Declock {#declock .unnumbered .hiddenheading}

My website provides many examples of the [Quarto](https://quarto.org) publishing and the [Dec](/dec) measurement systems in action. I leverage Quarto support for the [Observable](https://observablehq.com/) data analysis and visualization system to create animated and interactive graphics like the [bar📊charts](https://en.wikipedia.org/wiki/Bar_chart) clocks🕓, [solar☀️terminator](https://en.wikipedia.org/wiki/Terminator_%28solar%29#:~:text=a%20moving%20line%20that%20divides%20the%20daylit%20side%20and%20the%20dark%20night%20side%20of%20a%20planetary%20body) map🗺, [Earth🌍orbit](https://en.wikipedia.org/wiki/Earth%27s_orbit) diagram, and [daylight](https://en.wikipedia.org/wiki/Daylight#:~:text=the%20combination%20of%20all%20direct%20and%20indirect%20sunlight%20during%20the%20daytime) [area chart](https://en.wikipedia.org/wiki/Area_chart#:~:text=The%20area%20between%20axis%20and%20line%20are%20commonly%20emphasized%20with%20colors) below.

::: {.column-margin .clocks #onlymargin .posneg}
\${posclock}\${negclock}
:::

Dec times are measured in [fractional days](https://en.wikipedia.org/wiki/Decimal_time#Fractional_days:~:text=a%20decimal%20fraction%20of%20a%20day), often using [metric prefixes](https://en.wikipedia.org/wiki/Metric_prefix#:~:text=a%20unit%20prefix%20that%20precedes%20a%20basic%20unit%20of%20measure%20to%20indicate%20a%20multiple%20or%20submultiple%20of%20the%20unit) like [deci](https://en.wikipedia.org/wiki/Deci-#:~:text=a%20decimal%20unit%20prefix%20in%20the%20metric%20system%20denoting%20a%20factor%20of%20one%20tenth), [centi](https://en.wikipedia.org/wiki/Centi-#:~:text=a%20unit%20prefix%20in%20the%20metric%20system%20denoting%20a%20factor%20of%20one%20hundredth), or [milli](https://en.wikipedia.org/wiki/Milli-#:~:text=a%20unit%20prefix%20in%20the%20metric%20system%20denoting%20a%20factor%20of%20one%20thousandth). The top, middle, and bottom bars📊indicate the [decidays](https://en.wikipedia.org/wiki/Decimal_time#:~:text=dividing%20the%20day%20into%2010%20decidays){.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"}, [millidays](https://en.wikipedia.org/wiki/Unit_of_time#:~:text=1.44%20minutes%2C%20or%2086.4%20seconds){.tool data-bs-toggle="tooltip" data-bs-title="thousandths of a day"}, and [centimillidays](https://en.wikipedia.org/wiki/Decimal_time#:~:text=which%20would%20be-,1/100%2C000%20day,-%2C%20or%20864%20milliseconds){.tool data-bs-toggle="tooltip" data-bs-title="hundred thousandths of a day"}, respectively, of the time since the start, [+\${decTime}]{.point1}, or until the end, [-\${decTimeN}]{.point0}, of the day in the Dec [time zone](https://en.wikipedia.org/wiki/Time_zone#:~:text=an%20area%20which%20observes%20a%20uniform%20standard%20time), \${decZoneHsl}, at the location of the red⭕️circle on the map🗺️beneath the bars📊.

To rotate the globe🌐in the Earth🌏orbit diagram, drag the red⭕️circle horizontally↔️or slide the red🔴dot on the daylight area chart vertically↕. The red⭕️circle's vertical↕position determines the yearly daylight pattern visualized by the area chart. Shift the red<font color=red>—</font>line on the area chart horizontally↔️to move the globe🌐along the ellipse of the Earth🌎orbit.

{{< include /dec/time/_timecharts.qmd >}}

# Annual day aggregate (ada) {#sec-ada .hiddenheading}

The red<font color=red>—</font>line indicates a "day of year" ([doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"}), \${selDateHsl}, and the red🔴dot denotes a "time of day" ([tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}): \${selTimeDay}. A [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} identifies a day in a year like a [Gregorian calendar](https://en.wikipedia.org/wiki/Gregorian_calendar#:~:text=the%20calendar%20used%20in%20most%20parts%20of%20the%20world) month and "day of month" ([dom]{.tool data-bs-toggle="tooltip" data-bs-title="day of month"}). A [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} specifies a point in a day like an "hour minute second" ([hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"}) triplet.
Together, a [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} and [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} can form a "annual day aggregate" ([ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"}): \${selAdaDay}.

::: {.equationgroup #equationgroup00}
$$\begin{split}
\text{ada\;\,} & = \text{doy} + \text{tod} \\
\lfloor\text{ada}\rfloor & = \text{doy} \\
\end{split}$$ {#eq-ada}
:::

As their names suggest, [doy]{.tool data-bs-toggle="tooltip" data-bs-title="days of year"} and [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregates"} are measured in days. The [measurement unit](https://en.wikipedia.org/wiki/Unit_of_measurement#:~:text=a%20definite%20magnitude%20of%20a%20quantity) of a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} can be a day or a [submultiple](https://en.wikipedia.org/wiki/Multiple_%28mathematics%29#Submultiple:~:text=of%20%22a%20being-,a%20unit%20fraction,-of%20b%22%20) of a day. By changing how a decimal [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is measured, we can shift its [decimal separator](https://en.wikipedia.org/wiki/Decimal_separator#:~:text=a%20symbol%20that%20separates%20the%20integer%20part%20from%20the%20fractional%20part%20of%20a%20number) or turn it into an integer. The [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} along the y-axis of the area chart are integers because they have three digits and are measured in [millidays]{.tool data-bs-toggle="tooltip" data-bs-title="thousandths of a day"}.

# Epochal day aggregate (eda) {#sec-eda .hiddenheading}

To obtain a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} from a [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"}, we can [keep the remainder after dividing](https://en.wikipedia.org/wiki/Modulo#:~:text=returns%20the%20remainder) by one to isolate the [decimal part](https://en.wikipedia.org/wiki/Fractional_part#:~:text=the%20excess%20beyond%20that%20number%27s%20integer%20part) of the [quotient](https://en.wikipedia.org/wiki/Quotient#:~:text=a%20quantity%20produced%20by%20the%20division%20of%20two%20numbers): \${zeroAdaHsl} [mod](https://en.wikipedia.org/wiki/Modulo#:~:text=returns%20the%20remainder) 1 = \${zeroTimeHsl0}. We can use this same approach to separate a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} from an "epochal day aggregate" ([eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"}): \${zeroDoteHsl} mod 1 = \${zeroTimeHsl1}. The current [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} tells us how many days have passed since the Dec [epoch](https://en.wikipedia.org/wiki/Epoch#:~:text=an%20instant%20in%20time%20chosen%20as%20the%20origin%20of%20a%20particular%20calendar%20era).

::: {.equationgroup #equationgroup01}
$$\begin{split}
\text{tod} &= \text{ada} \bmod  1 &&= \text{eda} \bmod  1 \\
\text{tod} &= \text{ada} - \text{doy} &&= \text{eda} - \lfloor\text{eda}\rfloor
\end{split}$$ {#eq-todmod}
:::

When provided with an [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} instead of a [doe]{.tool data-bs-toggle="tooltip" data-bs-title="day of era"}, the Dec date equations return an [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} instead of a [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"}. While [doe]{.tool data-bs-toggle="tooltip" data-bs-title="days of era"} and [doy]{.tool data-bs-toggle="tooltip" data-bs-title="days of year"} are integers, [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregates"} and [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregates"} each have a decimal part called a "time of day" ([tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}). We can obtain an [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} by passing a [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"} and [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} to the Dec eda equations or by summing a [doe]{.tool data-bs-toggle="tooltip" data-bs-title="day of era"} and a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}.

::: {.equationgroup #equationgroup02}
$$\text{eda} = \text{coe}\times146097 + \text{yoc}\times365 + \lfloor\frac{\text{yoc}}{4}\rfloor - \lfloor\frac{\text{yoc}}{100}\rfloor + \text{ada}$$ {#eq-eda}

$$\text{tod} = \text{eda} \bmod  1 = \text{ada} \bmod  1$$ {#eq-tod}

$$\text{doe} = \lfloor \text{eda} \rfloor = \text{eda} - \text{tod}$$ {#eq-doe}

$$\text{doy} = \lfloor \text{ada} \rfloor = \text{ada} - \text{tod}$$ {#eq-doy}

$$\text{eda} = \text{doe} + \text{tod}$$ {#eq-edasum}

$$\text{ada} = \text{doy} + \text{tod}$$ {#eq-adasum}
:::

## UNIX time equation {#sec-ute .unnumbered .hiddenheading}

Similarly, [UNIX time](https://en.wikipedia.org/wiki/Unix_time#:~:text=the%20number%20of%20non%2Dleap%20seconds%20that%20have%20elapsed%20since%2000%3A00%3A00%20UTC%20on%201%C2%A0January%201970%2C%20the%20Unix%20epoch) tallies the seconds since the [UNIX epoch](https://en.wikipedia.org/wiki/Unix_time#:~:text=00%3A00%3A00%20UTC%20on%201%C2%A0January%201970), which is exactly 719468 days after the Dec epoch. To get the [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} in Zone [0]{.color0}, the Dec time zone that is in between the two leftmost vertical lines on the map🗺️, we can divide UNIX time by the [number of seconds in a day](https://en.wikipedia.org/wiki/Day#:~:text=average%2C%20this%20is-,24%20hours%20(86%2C400%20seconds),-.%20As%20a%20day) and then keep the remainder after dividing the resulting days by one:

::: {.equationgroup #equationgroup03}
$$\text{tod} = \text{unix} \div 86400 \bmod 1$$ {#eq-unixtod}
:::

## Julian date equation {#sec-jde .unnumbered .hiddenheading}

[Julian dates](https://en.wikipedia.org/wiki/Julian_day#:~:text=the%20Julian%20day%20number%20plus%20the%20fraction%20of%20a%20day%20since%20the%20preceding%20noon) track the days since the beginning of the [Julian period](https://en.wikipedia.org/wiki/Julian_day#:~:text=a%20chronological%20interval%20of%207980%C2%A0years%2C%20derived%20from%20three%20multi%2Dyear%20cycles:%20the%20Indiction%2C%20Solar%2C%20and%20Lunar%20cycles) and thus are akin to [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregates"}. We can produce a Zone [5]{.color5} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} from a [Julian date](https://en.wikipedia.org/wiki/Julian_day#:~:text=the%20Julian%20day%20number%20plus%20the%20fraction%20of%20a%20day%20since%20the%20preceding%20noon) simply by keeping the remainder after dividing by one. If we want a Zone [0]{.color0} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} instead, we should add [5]{.color5} [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} to the Julian date before converting it to a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} to ensure that the final result is less than one day:

::: {.equationgroup #equationgroup04}
$$\text{tod} = (\text{julian} + 0.5) \bmod 1$$ {#eq-juliantod}
:::

# Hour minute second {#sec-hms .hiddenheading}

We can also obtain a Zone [0]{.color0} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} from a [Coordinated Universal Time](https://en.wikipedia.org/wiki/Coordinated_Universal_Time#:~:text=the%20primary%20time%20standard%20globally%20used%20to%20regulate%20clocks%20and%20time) ([UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"}) [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplet by summing its components after converting them to fractional days, as shown in the equation below. The [computer programming](https://en.wikipedia.org/wiki/Computer_programming#:~:text=the%20composition%20of%20sequences%20of%20instructions%2C%20called%20programs%2C%20that%20computers%20can%20follow%20to%20perform%20tasks) code in the [tabset panel](https://quarto.org/docs/interactive/layout.html#tabset-panel) beneath the equation compares [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} derived from [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} and UNIX time as Quarto was [rendering](https://quarto.org/docs/get-started/hello/rstudio.html#rendering:~:text=When%20rendering%2C%20Quarto%20generates%20a%20new%20file%20that%20contains%20selected%20text%2C%20code%2C%20and%20results) this webpage.

::: {.equationgroup #equationgroup05}
$$\text{tod} = \frac{\text{hour}}{24} + \frac{\text{minute}}{1440} + \frac{\text{second}}{86400}$$ {#eq-hmstod}
:::

:::{.panel-tabset group="language"}
### Julia

```{julia}
using Dates
hms = now(UTC)
hour(hms) / 24 +
minute(hms) / 1440 +
second(hms) / 86400 +
millisecond(hms) / 864e5
datetime2unix(hms) / 86400 % 1
```

### Observable JavaScript

```{ojs}
hms = new Date();
hms.getUTCHours() / 24 +
hms.getUTCMinutes() / 1440 +
hms.getUTCSeconds() / 86400 +
hms.getUTCMilliseconds() / 864e5;
hms.getTime() / 864e5 % 1;
```

### Python

```{python}
from datetime import datetime, timezone
hms = datetime.now(timezone.utc)
hms.hour / 24 + \
hms.minute / 1440 + \
hms.second / 86400 + \
hms.microsecond / 864e8
hms.timestamp() / 86400 % 1
```

### R

```{r}
hms <- as.POSIXlt(Sys.time(), tz = "UTC")
hms$hour / 24 +
hms$min / 1440 +
hms$sec / 86400
(as.numeric(as.POSIXct(hms)) / 86400) %% 1
```

:::

The equations below convert UNIX time or a Zone [0]{.color0} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} into the three components of an [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplet: the "hour of day" ([hod]{.tool data-bs-toggle="tooltip" data-bs-title="hour of day"}), "minute of hour" ([moh]{.tool data-bs-toggle="tooltip" data-bs-title="minute of hour"}), and "second of minute" ([som]{.tool data-bs-toggle="tooltip" data-bs-title="second of minute"}), using a "daily second aggregate" ([dsa]{.tool data-bs-toggle="tooltip" data-bs-title="daily second aggregate"}) and "hourly second aggregate" ([hsa]{.tool data-bs-toggle="tooltip" data-bs-title="hourly second aggregate"}). While both count seconds, [dsa]{.tool data-bs-toggle="tooltip" data-bs-title="daily second aggregates"} start at midnight and [hsa]{.tool data-bs-toggle="tooltip" data-bs-title="hourly second aggregates"} begin at the [top of the hour](https://en.wiktionary.org/wiki/top_of_the_hour).

::: {.equationgroup #equationgroup06}
$$\begin{split}
\text{dsa} & = \text{tod} \times 86400 = \text{unix} \bmod  86400 \\
\text{hsa} & = \text{dsa} \bmod  3600 \\
\text{hod} & = \lfloor \text{dsa} \div 3600 \rfloor \\
\text{moh} & = \lfloor \text{hsa} \div 60 \rfloor \\
\text{som} & = \lfloor \text{hsa} \bmod  60 \rfloor
\end{split}$$ {#eq-hms}
:::


:::{.panel-tabset group="language"}
### Julia

```{julia}
using Dates
dsa = datetime2unix(now(UTC)) / 86400 % 1 * 86400
hsa = dsa % 3600
map(x -> floor(Int, x), (dsa / 3600, hsa / 60, hsa % 60))
```

### Observable JavaScript

```{ojs}
dsa = Date.now() / 864e5 % 1 * 86400;
hsa = dsa % 3600;
[dsa / 3600, hsa / 60, hsa % 60].map(Math.floor)
```

### Python

```{python}
from datetime import datetime, timezone
dsa = datetime.now(timezone.utc).timestamp() / 86400 % 1 * 86400
hsa = dsa % 3600
tuple(map(int, [dsa // 3600, hsa // 60, hsa % 60 // 1]))
```

### R

```{r}
dsa <- (as.numeric(as.POSIXct(Sys.time())) / 86400) %% 1 * 86400
hsa <- dsa %% 3600
sapply(c(dsa %/% 3600, hsa %/% 60, hsa %% 60), as.integer)
```

:::

# Universal time offset {#sec-uto .hiddenheading}

The [Global Positioning System](https://en.wikipedia.org/wiki/Global_Positioning_System#:~:text=a%20satellite%2Dbased%20hyperbolic%20navigation%20system), [BeiDou](https://en.wikipedia.org/wiki/BeiDou#:~:text=a%20satellite%2Dbased%20radio%20navigation%20system), and [Galileo](https://en.wikipedia.org/wiki/Galileo_(satellite_navigation)#:~:text=a%20global%20navigation%20satellite%20system%20(GNSS)%20created%20by%20the%20European%20Union) [global navigation satellite systems](https://en.wikipedia.org/wiki/Satellite_navigation#:~:text=a%20satellite%20navigation%20system%20with%20global%20coverage) along with most --- if not all --- [programming languages](https://en.wikipedia.org/wiki/Programming_language#:~:text=a%20system%20of%20notation%20for%20writing%20source%20code%20such%20as%20used%20to%20produce%20a%20computer%20program) do not account for [leap seconds](https://en.wikipedia.org/wiki/Leap_second#:~:text=one%2Dsecond%20adjustment), which appears to be for the best given that leap seconds will be [abolished by 2035](https://en.wikipedia.org/wiki/Leap_second#:~:text=at%20the%2027th%20General%20Conference%20on%20Weights%20and%20Measures%2C%20it%20was%20decided%20to%20abandon%20the%20leap%20second%20by%20or%20before%202035). The goal of leap seconds is to keep [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} within [25/24](https://en.wikipedia.org/wiki/Leap_second#:~:text=to%20ensure%20that%20the%20difference%20between%20the%20UTC%20and%20UT1%20readings%20will%20never%20exceed%200.9%20seconds) [centimillidays]{.tool data-bs-toggle="tooltip" data-bs-title="hundred thousands of a day"} ([cmd]{.tool data-bs-toggle="tooltip" data-bs-title="centimillidays"}) of [Universal Time](https://en.wikipedia.org/wiki/Universal_Time#Versions:~:text=a%20time%20standard%20based%20on%20Earth%27s%20rotation) ([UT]{.tool data-bs-toggle="tooltip" data-bs-title="Universal Time"}).

Instead of leap seconds, Dec matches [UT]{.tool data-bs-toggle="tooltip" data-bs-title="Universal Time"} using a "[universal time](https://en.wikipedia.org/wiki/Universal_Time#:~:text=a%20time%20standard%20based%20on%20Earth%27s%20rotation) [offset](https://en.wikipedia.org/wiki/UTC_offset#:~:text=the%20difference%20in%20hours%20and%20minutes%20between%20Coordinated%20Universal%20Time%20(UTC)%20and%20the%20standard%20time%20at%20a%20particular%20place)" ([uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"}). With the [leap second insertion dates](https://en.wikipedia.org/wiki/Leap_second#:~:text=Announced%20leap%20seconds%20to%20date) provided by the [International Earth Rotation and Reference Systems Service](https://en.wikipedia.org/wiki/International_Earth_Rotation_and_Reference_Systems_Service#:~:text=the%20body%20responsible%20for%20maintaining%20global%20time%20and%20reference%20frame%20standards), we can approximate the [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"} that yields [UT]{.tool data-bs-toggle="tooltip" data-bs-title="Universal Time"} when added to the Zone [0]{.color0} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} on the [Dec date](/dec/date) chosen by the [range](https://observablehq.com/framework/inputs/range)🎚️inputs below: \${leapCountHsl} &div; 8640 = \${leapTzoHsl0}.

```{ojs}
//| echo: false
//| label: dateinput
//| class: slider
viewof leapSecondYear = Inputs.range([1970, 2020], {label: "Year", value: 1995, step: 1})
viewof leapSecondDate = Inputs.range([0, 365], {label: "Day", value: 182, step: 1})
```

# Rounded offset decimal {#sec-rod .hiddenheading}

Of the twenty eight [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offsets"} that can be shown in the equation above, one is an integer, one is a [terminating decimal](https://en.wikipedia.org/wiki/Repeating_decimal#:~:text=a%20finite%20number%20of%20nonzero%20digits), and the rest are [repeating decimals](https://en.wikipedia.org/wiki/Repeating_decimal#:~:text=the%20same%20sequence%20of%20digits%20is%20repeated%20forever). To express a repeating decimal [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"}, Dec uses an [irreducible fraction](https://en.wikipedia.org/wiki/Irreducible_fraction#:~:text=a%20fraction%20in%20which%20the%20numerator%20and%20denominator%20are%20integers%20that%20have%20no%20other%20common%20divisors%20than%201) that is called an "exact offset fraction" ([eof]{.tool data-bs-toggle="tooltip" data-bs-title="exact offset fraction"}) when by itself or a "[rounding error](https://en.wikipedia.org/wiki/Round-off_error#:~:text=the%20difference%20between%20the%20result%20produced%20by%20a%20given%20algorithm%20using%20exact%20arithmetic%20and%20the%20result%20produced%20by%20the%20same%20algorithm%20using%20finite%2Dprecision%2C%20rounded%20arithmetic) fraction" ([ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"}) if it follows a "rounded offset decimal" ([rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"}).

In the equation below, the [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"} is the minuend, the [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"} is the subtrahend, and the [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"} is the difference. Dec uses the term minuend expansion to describe the replacement of a minuend with a subtrahend and a difference. By replacing a repeating decimal [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"} with a [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"} and a [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"}, we can show the initial digits of the [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"} as a decimal and the rest as a fraction.

::: {.equationgroup #equationgroup07}
$$\text{uto} - \text{rod} = \text{ref}$$ {#eq-ref}
:::

Use the first three range🎚️inputs below to select an [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplet to be converted to [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"}, plugged into the equation above as the [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"}, rounded to the number of digits chosen by the fourth range🎚️input, and inserted into the equation as the [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"}. Once the left-hand side of the equation is complete, we can solve it to get the [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"}: \${hmsTzoHsl} -- \${hmsRodHsl} = \${hmsRefHsl}.

```{ojs}
//| echo: false
//| label: hmsinput
//| class: slider
viewof hmsinput = Inputs.form([
  Inputs.range([0, 23], {label: "Hour", value: 11, step: 1}),
  Inputs.range([0, 59], {label: "Minute", value: 0, step: 1}),
  Inputs.range([0, 59], {label: "Second", value: 0, step: 1}),
])
```

```{ojs}
//| echo: false
//| label: diginput
//| class: slider
viewof digits = Inputs.range([1, 9], {label: "Digits", value: 3, step: 1})
```

# Time zone offset {#sec-tzo .hiddenheading}

In Dec, a [uto]{.tool data-bs-toggle="tooltip" data-bs-title="universal time offset"} can be any type of number, a "time zone offset" ([tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"}) is an integer, a "solar time offset" ([sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offset"}) is a terminating decimal, and an [eof]{.tool data-bs-toggle="tooltip" data-bs-title="exact offset fraction"} is a repeating decimal. If we derived [deciday]{.tool data-bs-toggle="tooltip" data-bs-title="a tenth of a day"} offsets from all 86400 of the [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplets that can be selected by the range🎚️inputs above, we would have 10 [tzos]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"}, 3190 [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offsets"}, and 83200 [eof]{.tool data-bs-toggle="tooltip" data-bs-title="exact offset fractions"} or [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"} and [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"} pairs.

## Coordinated Universal Time (UTC) {#sec-utc .unnumbered .hiddenheading}

When we do the same to the [38](https://en.wikipedia.org/wiki/List_of_UTC_offsets) [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} offsets, we get only [eof]{.tool data-bs-toggle="tooltip" data-bs-title="exact offset fractions"} or [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"} and [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"} pairs unless the number of leaps seconds included is zero or a multiple of 27. If the leap second count is zero or a multiple of 8640, we will get 3 [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"}, 9 [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offsets"}, and 26 [eof]{.tool data-bs-toggle="tooltip" data-bs-title="exact offset fractions"} or [rod]{.tool data-bs-toggle="tooltip" data-bs-title="rounded offset decimal"} and [ref]{.tool data-bs-toggle="tooltip" data-bs-title="roundoff error fraction"} pairs. The 3 [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"} will be [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offsets"} if the number of leap seconds is a multiple of 27 but not 8640.

There are 14 negative and 24 positive [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} offsets. The [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} time zone with [the most negative offset](https://en.wikipedia.org/wiki/UTC%E2%88%9212:00#:~:text=a%20nautical%20time%20zone%20comprising%20the%20high%20seas%20between%20180%C2%B0%20and%20172%C2%B030%E2%80%B2W%20longitude) is completely uninhabited. The bar📊chart below visualizes [Socioeconomic Data and Applications Center](https://www.earthdata.nasa.gov/centers/sedac-daac) data from 2020 regarding the population of each [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} time zone. The vast majority of all people live in [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} time zones with positive offsets.

{{< include /dec/time/_tzpopplot.qmd >}}

Negative [UTC]{.tool data-bs-toggle="tooltip" data-bs-title="Coordinated Universal Time"} offsets only exist in [the Americas](https://en.wikipedia.org/wiki/Americas#:~:text=a%20landmass%20comprising%20the%20totality%20of%20North%20America%20and%20South%20America) and islands in the [Atlantic](https://en.wikipedia.org/wiki/List_of_islands_in_the_Atlantic_Ocean) and [Pacific](https://en.wikipedia.org/wiki/List_of_islands_in_the_Pacific_Ocean) Oceans. Therefore, the bar📊chart above is essentially comparing the Americas to the rest of the world. According to 2021 [United Nations Department of Economic and Social Affairs](https://en.wikipedia.org/wiki/United_Nations_Department_of_Economic_and_Social_Affairs#:~:text=part%20of%20the%20United%20Nations%20Secretariat) data, about one billion out of a total of almost eight billion people live in the Americas.

Whenever a negative offset is associated with a Dec date, a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}, or both a date and a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}, Dec will add one day to the date and ten [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} to the offset without modifying the [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}. This typically occurs after the conversion of an [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplet to a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} or a "year month day" ([ymd]{.tool data-bs-toggle="tooltip" data-bs-title="year month day"}) triplet to a Dec date. As a result, all Dec dates and [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} have positive offsets.

Dec will not change a negative offset or its associated [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} if the result of adding one day to the [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} is uncertain. This uncertainly can only exist if we do not know whether a [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} that is equal to 364 belongs to a common or leap year. The day after Day [364]{.tool data-bs-toggle="tooltip" data-bs-title="February 28"} of a common year is Day [0]{.tool data-bs-toggle="tooltip" data-bs-title="March 1"} of the subsequent year. In a leap year, Day [364]{.tool data-bs-toggle="tooltip" data-bs-title="February 28"} precedes Day [365]{.tool data-bs-toggle="tooltip" data-bs-title="February 29"}.

# Day of week {#sec-dow .hiddenheading}

Even though it has no effect on the [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}, adding one day to the [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} also increments the "day of month" ([dom]{.tool data-bs-toggle="tooltip" data-bs-title="day of month"}) and "day of week" ([dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"}) shown by Dec. The table below shows how someone accustomed to a negative offset could intrepret Dec [dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"} numbers. From the perspective of a negative offset user, the [dom]{.tool data-bs-toggle="tooltip" data-bs-title="day of month"} and [dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"} in Dec will be one day ahead.

::: {.overflowtable}
| Saturday | Sunday | Monday | Tuesday | Wednesday | Thursday | Friday |
| -------- | ------ | ------ | ------- | --------- | -------- | ------ |
|    0     |   1    |   2    |    3    |     4     |    5     |   6    |
|   -7     |  -6    |  -5    |   -4    |    -3     |   -2     |  -1    |
::: {#tbl-dow}

The one day difference between positive and negative offsets may make Dec [dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"} numbers more intuitive than [POSIX](https://pubs.opengroup.org/onlinepubs/007904875/utilities/date.html#:~:text=weekday%20as%20a%20decimal%20number%20%5B0%2C6%5D%20(0%3Dsunday)) [dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"} numbers for people who consider [Sunday to be the first [dow]{.tool data-bs-toggle="tooltip" data-bs-title="day of week"}](https://en.wikipedia.org/wiki/Names_of_the_days_of_the_week#:~:text=Sunday%20comes%20first,-in%20order%20in). According to the [Common Locale Data Repository](https://en.wikipedia.org/wiki/Week#:~:text=World%20map%20showing%20the%20first%20day%20of%20the%20week%20used%20in%20different%20countries) and a [2023 population ranking](https://en.wikipedia.org/w/index.php?title=List_of_countries_in_the_Americas_by_population&oldid=1310903194#Table), over 89% of people in the Americas live in a country that starts the week on Sunday.

# Longitude and offsets {#sec-lao .hiddenheading}

In Dec, offsets are closely related to [longitude](https://en.wikipedia.org/wiki/Longitude#:~:text=denoted%20by%20the%20Greek%20letter%20lambda). Dec measures longitude in [parallels](https://en.wikipedia.org/wiki/Circle_of_latitude#:~:text=an%20abstract%20east%E2%80%93west%20small%20circle%20connecting%20all%20locations%20around%20Earth%20(ignoring%20elevation)%20at%20a%20given%20latitude%20coordinate%20line) ([&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="parallel"}) or submultiples of [&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="parallels"} like [deci[parallels](https://en.wikipedia.org/wiki/Circle_of_latitude#:~:text=an%20abstract%20east%E2%80%93west%20small%20circle%20connecting%20all%20locations%20around%20Earth%20(ignoring%20elevation)%20at%20a%20given%20latitude%20coordinate%20line)]{.tool data-bs-toggle="tooltip" data-bs-title="a tenth of a parallel"} ([d&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="deciparallels"}). A [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} is essentially a [d&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="deciparallel"} longitude that had its decimal part removed via [rounding](https://en.wikipedia.org/wiki/Rounding#:~:text=the%20process%20of%20adjusting%20a%20number%20to%20an%20approximate%2C%20more%20convenient%20value), [flooring](https://en.wikipedia.org/wiki/Floor_and_ceiling_functions#:~:text=takes%20as%20input%20a%20real%20number%20x%2C%20and%20gives%20as%20output%20the%20greatest%20integer%20less%20than%20or%20equal%20to%20x), [truncation](https://en.wikipedia.org/wiki/Truncation#Truncation_and_floor_function:~:text=truncation%20is%20limiting%20the%20number%20of%20digits%20right%20of%20the%20decimal%20point), or [ceiling](https://en.wikipedia.org/wiki/Floor_and_ceiling_functions#:~:text=maps%20x%20to%20the%20least%20integer%20greater%20than%20or%20equal%20to%20x). Whereas [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"} have one digit, [deciday]{.tool data-bs-toggle="tooltip" data-bs-title="a tenth of a day"} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} and [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offsets"} typically have up to four digits after the [decimal separator](https://en.wikipedia.org/wiki/Decimal_separator#:~:text=a%20symbol%20that%20separates%20the%20integer%20part%20from%20the%20fractional%20part%20of%20a%20number).

The fourth digit in the decimal part of any current deciday [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} increments [10^5^]{.tool data-bs-toggle="tooltip" data-bs-title="a hundred thousand"} times per day, 100 times per [milliday]{.tool data-bs-toggle="tooltip" data-bs-title="a thousandth of a day"}, or once per [beat]{.tool data-bs-toggle="tooltip" data-bs-title="a hundred thousandth of a day"} ([iob]{.tool data-bs-toggle="tooltip" data-bs-title="inverse of beat"}), which is the [lower bound](https://en.wikipedia.org/wiki/Upper_and_lower_bounds#:~:text=an%20element%20of%20K%20that%20is%20less%20than%20or%20equal%20to%20every%20element%20of%20S) of the [normal resting heart rate](https://en.wikipedia.org/wiki/Heart_rate#:~:text=normal%20resting%20adult%20human%20heart%20rate%20is%2060–100%20bpm) of an adult.
For [everyday life](https://en.wikipedia.org/wiki/Everyday_life), we should limit the length of [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} to the three digits needed to show [millidays]{.tool data-bs-toggle="tooltip" data-bs-title="thousandths of a day"} ([md]{.tool data-bs-toggle="tooltip" data-bs-title="millidays"}) or the five digits required to display [beats]{.tool data-bs-toggle="tooltip" data-bs-title="hundred thousandths of a day"} ([b]{.tool data-bs-toggle="tooltip" data-bs-title="beats"}).

When the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} has seven digits, the sixth digit changes too quickly to be read out loud and the seventh changes so fast that it appears as a blur. Near the Equator, a longitude that has seven digits is accurate to within about ten [zem]{.tool data-bs-toggle="tooltip" data-bs-title="zone equatorial meters"} (z) or four [meters](https://en.wikipedia.org/wiki/Metre#:~:text=the%20base%20unit%20of%20length%20in%20the%20International%20System%20of%20Units), which is roughly the length of a [subcompact car](https://www.dimensions.com/collection/subcompact-cars-b-segment) or the width of a [U-shaped living room layout](https://www.dimensions.com/collection/living-room-layouts).

The Equator is approximately ([~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}) [10^3^]{.tool data-bs-toggle="tooltip" data-bs-title="one thousand"} [millitaurs]{.tool data-bs-toggle="tooltip" data-bs-title="thousandths of a taur"} ([mc]{.tool data-bs-toggle="tooltip" data-bs-title="millitaurs"}), [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}4 &times; [10^4^]{.tool data-bs-toggle="tooltip" data-bs-title="ten thousand"} [kilometers]{.tool data-bs-toggle="tooltip" data-bs-title="thousands of meters"} ([km]{.tool data-bs-toggle="tooltip" data-bs-title="kilometers"}), or [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}[10^5^]{.tool data-bs-toggle="tooltip" data-bs-title="a hundred thousand"} [kilozem]{.tool data-bs-toggle="tooltip" data-bs-title="thousands of zone equatorial meters"} ([kz]{.tool data-bs-toggle="tooltip" data-bs-title="kilozems"}) long. If we move 1 [mc]{.tool data-bs-toggle="tooltip" data-bs-title="millitaurs"}, 40 [km]{.tool data-bs-toggle="tooltip" data-bs-title="kilometers"}, or 100 [kz]{.tool data-bs-toggle="tooltip" data-bs-title="kilozems"} to the East or West on or near the Equator, our [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offset"} will change by [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}1 [md]{.tool data-bs-toggle="tooltip" data-bs-title="millidays"}, [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}1.44 minutes, or [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}100 [b]{.tool data-bs-toggle="tooltip" data-bs-title="beat"} and our longitude will shift by [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}0.36 degrees, [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}1 milliparallel ([m&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="milliparallel"}), [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}21.6 [arcminutes](https://en.wikipedia.org/wiki/Minute_and_second_of_arc){.tool data-bs-toggle="tooltip" data-bs-title="a sixtieth of a degree"}, or [~]{.tool data-bs-toggle="tooltip" data-bs-title="approximately"}100 arcbeats ([ab]{.tool data-bs-toggle="tooltip" data-bs-title="arcbeats"}).

For precise [geopositioning](https://en.wikipedia.org/wiki/Geopositioning#:~:text=estimating%20the%20geographic%20position%20of%20an%20object%20or%20a%20person), it may be helpful to show [geographic coordinates](https://en.wikipedia.org/wiki/Geographic_coordinate_system#:~:text=for%20measuring%20and%20communicating%20positions%20directly%20on%20Earth%20as%20latitude%20and%20longitude) in submultiples of [ab]{.tool data-bs-toggle="tooltip" data-bs-title="arcbeats"}, but we are unlikely to benefit from units smaller than [md]{.tool data-bs-toggle="tooltip" data-bs-title="millidays"} when displaying the [solar time](https://en.wikipedia.org/wiki/Solar_time#:~:text=a%20calculation%20of%20the%20passage%20of%20time%20based%20on%20the%20position%20of%20the%20Sun%20in%20the%20sky) or estimates of what the [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} will be when the Sun rises, reaches its [zenith](https://en.m.wikipedia.org/wiki/Noon#:~:text=highest%20position%20above%20the%20horizon), or sets on a given day. By default, Dec uses three digits to show each solar time and [sto]{.tool data-bs-toggle="tooltip" data-bs-title="solar time offset"}.

# Equation of time {#sec-eot .hiddenheading}

The two types of [solar time](https://en.wikipedia.org/wiki/Solar_time#:~:text=a%20calculation%20of%20the%20passage%20of%20time%20based%20on%20the%20position%20of%20the%20Sun%20in%20the%20sky) are ["mean solar time"](https://en.wikipedia.org/wiki/Solar_time#Mean_solar_time:~:text=it%20follows%20an%20imaginary%20%22mean%20Sun%22%20that%20moves%20along%20the%20celestial%20equator%20at%20a%20constant%20rate%20that%20matches%20the%20real%20Sun%27s%20average%20rate%20over%20the%20year) ([mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"}) and ["apparent solar time"](https://en.wikipedia.org/wiki/Solar_time#Mean_solar_time:~:text=it%20follows%20an%20imaginary%20%22mean%20Sun%22%20that%20moves%20along%20the%20celestial%20equator%20at%20a%20constant%20rate%20that%20matches%20the%20real%20Sun%27s%20average%20rate%20over%20the%20year) ([ast]{.tool data-bs-toggle="tooltip" data-bs-title="apparent solar time"}). To calculate [mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"}, we keep only the decimal part of the sum of 0.95, the Zone [0]{.color0} [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} measured in days, and our longitude measured in [&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="parallel"}. If we want [ast]{.tool data-bs-toggle="tooltip" data-bs-title="apparent solar time"} instead of [mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"}, the sum needs to include the result of plugging the "time of year" ([toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}) into the ["equation of time"](https://en.wikipedia.org/wiki/Equation_of_time#:~:text=the%20discrepancy%20between%20two%20kinds%20of%20solar%20time) ([eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"}).

::: {.equationgroup #equationgroup08}
$$\text{toy} = \text{ada} \div \text{n}$$ {#eq-toy}

$$\text{mst} = (0.95 + \text{tod} + \lambda) \bmod 1$$ {#eq-mst}

$$\text{ast} = (0.95 + \text{tod} + \lambda +  \text{eot(toy)}) \bmod 1$$ {#eq-ast}
:::

To obtain the [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}, we divide the [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} by the number of days in the year ([n]{.tool data-bs-toggle="tooltip" data-bs-title="the number of days in the year"}). If we use [trigonometric functions](https://en.wikipedia.org/wiki/Trigonometric_functions#:~:text=functions%20which%20relate%20an%20angle%20of%20a%20right%2Dangled%20triangle%20to%20ratios%20of%20two%20side%20lengths) that are designed to work with [radians](https://en.wikipedia.org/wiki/Radian#:~:text=the%20unit%20of%20angle%20in%20the%20International%20System%20of%20Units), we will have to multiply the [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"} by $2\pi$ or $\tau$. We do not need to modify the [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"} before passing it to the [eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"}() function defined below because its trigonometric functions expect [turns](https://en.wikipedia.org/wiki/Turn_%28angle%29#:~:text=a%20unit%20of%20plane%20angle%20measurement%20equal%20to%202%CF%80%C2%A0radians%2C%20360%C2%A0degrees) instead of radians.

::: {.equationgroup #equationgroup09}
$$\begin{split}
\text{eot(toy)} & = \beta_0 \\
& + \beta_1 \times \text{costau(toy)} \\
& + \beta_2 \times \text{costau(2} \times \text{toy)} \\
& + \beta_3 \times \text{sintau(toy)} \\
& + \beta_4 \times \text{sintau(2} \times \text{toy)}
\end{split}$$ {#eq-eot}
:::

:::{.panel-tabset group="language"}
### Julia

```{julia}
sincostau(x) = sincospi(2 * x)
sincostau(0.25)
sincostau(0.5)
sincos(0.25 * 2 * pi)
sincos(0.5 * 2 * pi)
```

### Observable JavaScript

```{ojs}
sincospi = require(
  'https://cdn.jsdelivr.net/gh/stdlib-js/math-base-special-sincospi@umd/browser.js'
  )
function sincostau(x) {
  return sincospi(2 * x);
}
function sincos(x) {
  x *= 2 * Math.PI
  return [Math.sin(x), Math.cos(x)]
}
sincostau(0.25)
sincostau(0.5)
sincos(0.25)
sincos(0.5)
```

### Python

```{python}
from math import cos, sin, pi
from mpmath import cospi, sinpi

def sincostau(x):
    x *= 2
    return sinpi(x), cospi(x)

def sincos(x):
    x *= 2 * pi
    return sin(x), cos(x)

sincostau(0.25)
sincostau(0.5)
sincos(0.25)
sincos(0.5)
```

### R

```{r}
sincostau <- function(x) {
  x <- 2 * x
  c(sinpi(x), cospi(x))
}

sincos <- function(x) {
  x <- 2 * x * pi
  c(sin(x), cos(x))
}

sincostau(0.25)
sincostau(0.5)
sincos(0.25)
sincos(0.5)
```

:::

The code above compares two functions which return both the [sine](https://en.wikipedia.org/wiki/Sine_and_cosine#:~:text=the%20ratio%20of%20the%20length%20of%20the%20side%20opposite%20that%20angle%20to%20the%20length%20of%20the%20longest%20side%20of%20the%20triangle) and [cosine](https://en.wikipedia.org/wiki/Sine_and_cosine#:~:text=the%20ratio%20of%20the%20length%20of%20the%20adjacent%20leg%20to%20that%20of%20the%20hypotenuse) of their input. The sincos() function assumes it will receive radians and the sincostau() works with turns. In general, working with turns is more convenient and intuitive. The examples above demonstrate that using turns instead of radians can [yield more accurate results](https://en.wikipedia.org/wiki/Sine_and_cosine#:~:text=Representing%20angles%20in%20turns%20or%20half%2Dturns%20has%20accuracy%20advantages).

The values below are the coefficients of a model adapted from the [National Oceanic and Atmospheric Administration](https://www.noaa.gov) ([NOAA]{.tool data-bs-toggle="tooltip" data-bs-title="National Oceanic and Atmospheric Administration"}) [General Solar Position Calculations](https://gml.noaa.gov/grad/solcalc/solareqns.PDF). Before [fitting](https://docs.scipy.org/doc/scipy/reference/generated/scipy.optimize.curve_fit.html) the model to [NOAA]{.tool data-bs-toggle="tooltip" data-bs-title="National Oceanic and Atmospheric Administration"} [yearly solar data](https://gml.noaa.gov/grad/solcalc/calcdetails.html), we need to convert eot([toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}) values from minutes to [centidays](https://en.wikipedia.org/wiki/Unit_of_time#:~:text=14.4%20minutes%2C%20or%20864%20seconds){.tool data-bs-toggle="tooltip" data-bs-title="hundredths of a day"}, combine [ymd]{.tool data-bs-toggle="tooltip" data-bs-title="year month day"} triplet dates and [hms]{.tool data-bs-toggle="tooltip" data-bs-title="hour minute second"} triplet times into [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregates"}, and sort by [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"}.

{{< embed eot.ipynb#eotcoef >}}

{{< include /dec/time/_eotplot.qmd >}}

The [line📈chart](https://en.wikipedia.org/wiki/Line_chart#:~:text=a%20type%20of%20chart%20that%20displays%20information%20as%20a%20series%20of%20data%20points%20called%20%27markers%27%20connected%20by%20straight%20line%20segments) above uses [md]{.tool data-bs-toggle="tooltip" data-bs-title="millidays"} to display [eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"}([toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}) values as integers. There is little difference between [mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"} and [ast]{.tool data-bs-toggle="tooltip" data-bs-title="apparent solar time"} around Days [45]{.tool data-bs-toggle="tooltip" data-bs-title="April 15"}, [103]{.tool data-bs-toggle="tooltip" data-bs-title="June 12"}, [184]{.tool data-bs-toggle="tooltip" data-bs-title="September 1"}, and [299]{.tool data-bs-toggle="tooltip" data-bs-title="December 25"}. The difference ranges from about -9.8 on [Day 244]{.tool data-bs-toggle="tooltip" data-bs-title="October 31"} to around 11.4 [md]{.tool data-bs-toggle="tooltip" data-bs-title="millidays"} on [Day 350]{.tool data-bs-toggle="tooltip" data-bs-title="February 14"}. We can calculate [mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"} with just a [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and a longitude and we will be off by at most about a [centiday]{.tool data-bs-toggle="tooltip" data-bs-title="a hundredth of a day"} compared to [ast]{.tool data-bs-toggle="tooltip" data-bs-title="apparent solar time"}.

Apart from turning a [mst]{.tool data-bs-toggle="tooltip" data-bs-title="mean solar time"} into an [ast]{.tool data-bs-toggle="tooltip" data-bs-title="apparent solar time"}, we can also use [eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"} to more accurately estimate the [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} of [solar noon](https://en.wikipedia.org/wiki/Noon#:~:text=reaching%20its%20highest%20position%20above%20the%20horizon%20on%20that%20day%20and%20casting%20the%20shortest%20shadow), sunrise, and sunset. The equation below creates a [solar noon](https://en.wikipedia.org/wiki/Noon#:~:text=reaching%20its%20highest%20position%20above%20the%20horizon%20on%20that%20day%20and%20casting%20the%20shortest%20shadow) [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} measured in days by adding 9.55 to a longitude measured in [&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="parallel"}, subtracting a [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} and a [eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"}([toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}) value that are both measured in days, and keeping only the decimal part of the result.

::: {.equationgroup #equationgroup10}
$$\text{solarnoon} = \left(9.55 + \text{tzo} - \lambda - \text{eot(toy)}\right) \bmod 1$$ {#eq-solarnoon}
:::

# Cambridge and Cambridge {#sec-cac .hiddenheading}

To compare the solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} in two cities, we can plug in the longitude of each city into the equation above. If all the variables in the equation other than longitude are set to zero, the result is almost five [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} for [Cambridge, England](https://en.wikipedia.org/wiki/Cambridge#:~:text=a%20city%20and%20non%2Dmetropolitan%20district%20in%20the%20county%20of%20Cambridgeshire%2C%20England) in the [United Kingdom](https://en.wikipedia.org/wiki/United_Kingdom#:~:text=a%20country%20in%20Northwestern%20Europe) and nearly seven [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} for [Cambridge, Massachusetts](https://en.wikipedia.org/wiki/Cambridge,_Massachusetts#:~:text=a%20city%20in%20Middlesex%20County%2C%20Massachusetts%2C%20United%20States) in the [United States](https://en.wikipedia.org/wiki/United_States#:~:text=a%20country%20primarily%20located%20in%20North%20America).

::: {.equationgroup #equationgroup11}
$$4.99635 = (9.55 - 0.050365) \bmod 1 \times 10$$ {#eq-cambridgeuk}

$$6.97516 = (9.55 - 0.852484) \bmod 1 \times 10$$ {#eq-cambridgema}
:::

The two homonymous cities are about two [d&lambda;]{.tool data-bs-toggle="tooltip" data-bs-title="deciparallels"} apart and thus will always differ by around two [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} in solar time, regardless of what time zone we use as our frame of reference. England is in Zone [0]{.color0} and Massachusetts is in Zone [8]{.color8}. If we change the [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} from zero to eight [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"}, the solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} for each city will be two [decidays]{.tool data-bs-toggle="tooltip" data-bs-title="tenths of a day"} earlier.

::: {.equationgroup #equationgroup12}
$$2.99635 = (9.55 + 0.8 - 0.050365) \bmod 1 \times 10$$ {#eq-cambridgeuk8}

$$4.97516 = (9.55 + 0.8 - 0.852484) \bmod 1 \times 10$$ {#eq-cambridgema8}
:::

# Full day arc {#sec-fda .hiddenheading}

The path in the sky that the Sun appears to follow from a sunrise to a sunset is a "[day arc](https://en.wikipedia.org/wiki/Sun_path#:~:text=arc%2Dlike%20path%20that%20the%20Sun%20appears%20to%20follow%20across%20the%20sky)" ([da]{.tool data-bs-toggle="tooltip" data-bs-title="full day arc"}). Solar noon, the midpoint of a [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"}, is halfway between sunrise and sunset. On each side of solar noon is a "half day arc" ([da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"}). The sum of a solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and a [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"} is a sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and the difference between a solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and a [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"} is a sunrise [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}.

::: {.equationgroup #equationgroup13}
$$\begin{split}
\text{sunset} & = \text{solarnoon} + \frac{\text{da}}{2} && = \text{sunrise} + \text{da} \\
\text{sunrise} & = \text{solarnoon} - \frac{\text{da}}{2} && = \text{sunset} - \text{da}
\end{split}$$ {#eq-sunsetsunrise}
:::

## Full night arc {#sec-fna .unnumbered .hiddenheading}

The Sun continues its path after it disappears below the horizon, moving from sunset to sunrise along a "night arc" ([na]{.tool data-bs-toggle="tooltip" data-bs-title="night arc"}). Like [conjugate angles](https://en.wikipedia.org/wiki/Angle#:~:text=conjugate%20angles%20sum%20to%20a%20full%20angle), a [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} and a [na]{.tool data-bs-toggle="tooltip" data-bs-title="night arc"} form a full circle that represents one day. Likewise, a [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"} and a "half night arc" ([na/2]{.tool data-bs-toggle="tooltip" data-bs-title="half night arc"}) are like two [supplementary angles](https://en.wikipedia.org/wiki/Angle#:~:text=sum%20to%20a%20straight%20angle) that form a semicircle from solar noon to the [na]{.tool data-bs-toggle="tooltip" data-bs-title="night arc"} midpoint: [solar midnight](https://en.wikipedia.org/wiki/Midnight#:~:text=the%20time%20opposite%20to%20solar%20noon%2C%20when%20the%20Sun%20is%20closest%20to%20the%20nadir%2C%20and%20the%20night%20is%20equidistant%20from%20dusk%20and%20dawn).

::: {.equationgroup #equationgroup14}
$$\text{solarmidnight} = \text{solarnoon} + \frac{\text{da}}{2} + \frac{\text{na}}{2} = \text{sunset} + \frac{\text{na}}{2}$$ {#eq-solarmidnight}
:::

The range input below controls the yellow [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} and the blue [na]{.tool data-bs-toggle="tooltip" data-bs-title="night arc"} in the diagram beneath it. At the Equator, the sunrise always rises about a quarter turn from North and sets around three quarter turns from North, resulting in a [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} of approximately half a day: 75% -- 25% = 50%. If [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} is zero, a [polar night](https://en.wikipedia.org/wiki/Polar_night#:~:text=the%20Sun%20remains%20below%20the%20horizon%20for%20more%20than%2024%C2%A0hours) occurs. A day without a [na]{.tool data-bs-toggle="tooltip" data-bs-title="night arc"} is called a [polar day](https://en.wikipedia.org/wiki/Midnight_sun#:~:text=the%20Sun%20remains%20visible%20at%20the%20local%20midnight).

{{< include /dec/time/_dayarcplot.qmd >}}

The diagram above can represent both a clock and a compass. In the [Northern Hemisphere](https://en.wikipedia.org/wiki/Northern_Hemisphere#:~:text=the%20half%20of%20Earth%20that%20is%20north%20of%20the%20equator), solar time is essentially the same as the ["solar azimuth angle"](https://en.wikipedia.org/wiki/Solar_azimuth_angle#:~:text=horizontal%20angle%20with%20respect%20to%20north) ([saa]{.tool data-bs-toggle="tooltip" data-bs-title="solar azimuth angle"}). In the [Southern Hemisphere](https://en.wikipedia.org/wiki/Southern_Hemisphere#:~:text=the%20half%20(hemisphere)%20of%20Earth%20that%20is%20south%20of%20the%20equator), we would need to flip the diagram upside down for it to match the [saa]{.tool data-bs-toggle="tooltip" data-bs-title="solar azimuth angle"}. The bar chart below compares the populations of the Northern and Southern Hemispheres.

{{< include /dec/time/_hemipopplot.qmd >}}

Anyone can measure the [saa]{.tool data-bs-toggle="tooltip" data-bs-title="solar azimuth angle"} during the daytime by pointing a compass at the point on the horizon below the Sun. In the Northern Hemisphere, we can use this method to approximate solar time. In the Southern Hemisphere, we can obtain solar time by measuring the [saa]{.tool data-bs-toggle="tooltip" data-bs-title="solar azimuth angle"} in turns and then subtracting our measurement from one and a half turns:

::: {.equationgroup #equationgroup15}
$$\text{solartime} \approx \begin{cases}\text{saa}&{\text{if } \phi \geq 0;}\\(1.5 - \text{saa}) \bmod 1&{\text{otherwise.}}\end{cases}$$ {#eq-solartime}
:::

The clockwise path that the Sun follows in the Northern Hemisphere is ingrained in the Belarussian, Polish, or Ukrainian languages. These three Slavic languages each have one word for north or midnight and another word for south or noon. In Dec, zero represents both north and midnight, while both south and noon can be expressed as a half turn.

::: {.overflowtable}
 | Turns | Direction | Time     | Belarussian | Polish   | Ukrainian |
 | ----- | --------- | -------- | ----------- | -------- | --------- |
 | 0     | north     | midnight | поўнач      | północ   | північ    |
 | 0.5   | south     | noon     | поўдзень    | południe | південь   |
::: {#tbl-slav}

If we only want to know how long the Sun will shine on a given day, we can use the top equation below to obtain a [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"}. Alternatively, if we are interested in finding out when the Sun will rise or set on a given day, we will need to calculate a [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"} using the bottom equation below and then combine it with a solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} to get a sunrise or sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}.


::: {.equationgroup #equationgroup16}
$$\text{da} = \frac{\arccos\left({\Large\frac{\text{costau(0.252314)} - \text{sintau($\phi$)} \times \text{sintau(sda)}}{\text{costau($\phi$)} \times \text{costau(sda)}}}\right)}{\pi}$$ {#eq-da}

$$\frac{\text{da}}{2} = \frac{\arccos\left({\Large\frac{\text{costau(0.252314)} - \text{sintau($\phi$)} \times \text{sintau(sda)}}{\text{costau($\phi$)} \times \text{costau(sda)}}}\right)}{\tau}$$ {#eq-halfda}
:::

## Solar declination angle {#sec-sda .unnumbered .hiddenheading}

The [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} and [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"} equations above require a latitude and a "solar [declination](https://en.wikipedia.org/wiki/Declination#:~:text=one%20of%20the%20two%20angles%20that%20locate%20a%20point%20on%20the%20celestial%20sphere%20in%20the%20equatorial%20coordinate%20system) angle" ([sda]{.tool data-bs-toggle="tooltip" data-bs-title="solar declination angle"}). Dec measures latitude in turns called meridians ([&phi;]{.tool data-bs-toggle="tooltip" data-bs-title="meridians"}) or turn submultiples such as millimeridians ([m&phi;]{.tool data-bs-toggle="tooltip" data-bs-title="millimeridians"}). For simplicity, we can fit our [eot]{.tool data-bs-toggle="tooltip" data-bs-title="equation of time"} model to [sda]{.tool data-bs-toggle="tooltip" data-bs-title="solar declination angle"} data instead of fitting the needlessly complex [sda]{.tool data-bs-toggle="tooltip" data-bs-title="solar declination angle"} model provided by the NOAA General Solar Position Calculations:

::: {.equationgroup #equationgroup17}
$$\begin{split}
\text{sda(toy)} & = \beta_0 \\
& + \beta_1 \times \text{costau(toy)} \\
& + \beta_2 \times \text{sintau(toy)} \\
& + \beta_3 \times \text{costau(2} \times \text{toy)} \\
& + \beta_4 \times \text{sintau(2} \times \text{toy)} \\
& + \beta_5 \times \text{costau(3} \times \text{toy)} \\
& + \beta_6 \times \text{sintau(3} \times \text{toy)} \\
\end{split}$$ {#eq-sda}
:::

The top range input below picks the [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} that will become the [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"} in our fitted [sda]{.tool data-bs-toggle="tooltip" data-bs-title="solar declination angle"} model. The other two range inputs below chose the geographic coordinates that we need to find the sunrise and sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"}. While the sunrise and sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} depend on both geographic coordinates, the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} varies only by latitude and not by longitude: ${selectedSunset.toFixed(3)} -- ${selectedSunrise.toFixed(3)} = \${parseFloat(daytimeDuration.toFixed(3))}.

```{ojs}
//| echo: false
//| label: doyinput
//| class: slider
viewof doyInput = Inputs.range([0, 365], {label: "Day of year", value: 183, step: 1})
```

```{ojs}
//| echo: false
//| label: latinput
//| class: slider
viewof latInput = Inputs.range([-250, 250], {label: "Latitude", value: 0, step: 1})
```

```{ojs}
//| echo: false
//| label: loninput
//| class: slider
viewof lonInput = Inputs.range([0, 1000], {label: "Longitude", value: 500, step: 1})
```

The equation controlled by range inputs above can be summarized as sunset -- sunrise = [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} and can be rearranged into the sunrise equation: sunrise = sunset -- [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"}. If we plug the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} in place of the sunrise [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} in the sunrise equation without changing the sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}, the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} will be replaced with the time until or since sunset: ${selectedCurrent.toFixed(3)} = ${selectedSunset.toFixed(3)} ${selectedDifference < 0 ? "--" : "+"} ${parseFloat(Math.abs(selectedDifference).toFixed(3))}.

The latest equation above is an example of a Dec span. From left to right, this span consists of a [minuend](https://en.wiktionary.org/wiki/minuend#:~:text=A%20number%20or%20quantity%20from%20which%20another%20is%20to%20be%20subtracted), an equals sign (=), [subtrahend](https://en.wikipedia.org/wiki/Subtraction#:~:text=number%20being%20subtracted), a plus (+) or a minus (--) sign, and a [difference](https://en.wikipedia.org/wiki/Subtraction#Notation_and_terminology:~:text=The%20result%20is%20the%20difference). A minus sign indicates the difference is the time remaining **until** sunset, whereas a plus sign means that the difference is the time that has passed **since** sunset.

The difference shows the [remaining time](https://en.wikipedia.org/wiki/Countdown#:~:text=the%20time%20remaining%20before%20an%20event) as a negative number if the minuend is less than the substrahend or displays the [elapsed time](https://en.wikipedia.org/wiki/Mission_Elapsed_Time) as a positive number if the minuend is greater than the subtrahend. In this way, the difference can act like a ["T minus countdown"](https://en.wikipedia.org/wiki/Countdown#:~:text=a%20sequence%20of%20backward%20counting%20to%20indicate%20the%20time%20remaining) that becomes a ["T plus countup"](https://en.wikipedia.org/wiki/Countdown#:~:text=After%20a%20launch%2C%20most%20countdown%20clocks%20begin%20to%20show%20Mission%20Elapsed%20Time) after an event like a rocket🚀launch.

::: {.equationgroup #equationgroup18}
$$\begin{cases}\text{difference}<0&{\text{if } \text{minuend} < \text{subtrahend};}\\\text{difference}=0&{\text{if minuend} = \text{subtrahend};}\\\text{difference}>0&{\text{if minuend}>\text{subtrahend.}}\end{cases}$$ {#eq-difference}
:::

When the minuend is the current time and the subtrahend is the [timestamp](https://en.wikipedia.org/wiki/Timestamp#:~:text=a%20sequence%20of%20characters%20or%20encoded%20information%20identifying%20when%20a%20certain%20event%20occurred) of an event which occurred in the past, the difference is the time elapsed since that event. A Dec timestamp consists of a year, a day, . can be an [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"}, a "epochal year aggregate" ([eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"}), or a snap🫰. Eda and eya are  the time The typical snap format is year[+]{.mono}[ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"}[-]{.mono}[tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"}. The snap \${loadYearHsl}[+]{.mono}\${loadAdaHsl}[-]{.mono}[0]{.color0} represents the [b]{.tool data-bs-toggle="tooltip" data-bs-title="beat"} when this webpage loaded ${elaTimeHsl} [b]{.tool data-bs-toggle="tooltip" data-bs-title="beats"} ago.

# Epochal year aggregate (eya) {#sec-eya .hiddenheading}

A [doe]{.tool data-bs-toggle="tooltip" data-bs-title="day of era"} is essentially a Dec date with a [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"} that is always equal to 0 and a [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} that is [not restricted](https://en.wikipedia.org/wiki/Bounded_set#:~:text=a%20set%20which%20is%20not%20bounded) to 0 &le; [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} &le; 365. Similarly, a Dec "epochal year aggregate" ([eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"}) is basically a Dec date with a non-integer year and a day permanently set to [0]{.color0}. We can obtain a Dec date by passing a doe Both [doe]{.tool data-bs-toggle="tooltip" data-bs-title="days of era"} and [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregates"} allow us to represent a date as a single number and obtain the difference between two dates.

::: {.equationgroup #equationgroup19}
$$\text{cda} = \text{de} - \text{coe} \times 146097$$ {#eq-cda}

$$\text{yoc} = \biggl \lfloor \frac{\text{doc} - \lfloor \frac{\text{doc}}{1460} \rfloor + \lfloor \frac{\text{doc}}{36524} \rfloor - \lfloor \frac{\text{doc}}{146096} \rfloor}{365} \biggr \rfloor$$ {#eq-yocdoc}

$$\text{yoe} = \text{yoc} + \text{coe} \times 400$$ {#eq-yoe}

$$\text{doy} = \biggl \lfloor \text{doc} - \text{yoc} \times 365 - \lfloor \frac{\text{yoc}}{4} \rfloor + \lfloor \frac{\text{yoc}}{100} \rfloor \biggr \rfloor$$ {#eq-doydoc}
:::

Compared to a [doe]{.tool data-bs-toggle="tooltip" data-bs-title="day of era"}, it is much easier to convert between an [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"} and a to turn into Dec dates. The decimal part of an [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"} is a called a [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}. We can convert between dates to [eya]{.tool data-bs-toggle="tooltip" data-bs-title="times of year"} with the equations below. We can obtain a [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"} by flooring an [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"}. Likewise, we can obtain a [doy]{.cyan .under .tool data-bs-toggle="tooltip" data-bs-title="day of year"} by flooring the product of [n]{.tool data-bs-toggle="tooltip" data-bs-title="number of days in the year"} and the decimal part of an [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"}, which is called a [toy]{.tool data-bs-toggle="tooltip" data-bs-title="time of year"}. The current [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"} equation values are \${fullfracYear} = \${decYear} + \${decDoty} &div; \${nDaysInYear}. We can floor an [eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"} to get a yoe or divide its , we can obtain a 

::: {.equationgroup #equationgroup20}
$$\text{eya} = \text{yoe} + \text{toy} = \text{yoe} + \text{doy} \div \text{n}$$ {#eq-eya}

$$\text{toy} = \text{eya} \bmod 1 = \text{doy} \div \text{n}$$ {#eq-toyeya}

$$\text{doy} = \lfloor\text{toy}\times\text{n}\rfloor$$ {#eq-doytoy}

$$\text{eya} = \text{yoe} + \text{doy} \times \text{n}$$ {#eq-eyadoy}

$$\text{eya} = \text{yoe} + \text{coe} \times 400$$ {#eq-eyacoe}
:::

[eya]{.tool data-bs-toggle="tooltip" data-bs-title="epochal year aggregate"} = [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"} + [doy]{.tool data-bs-toggle="tooltip" data-bs-title="day of year"} &div; [n]{.tool data-bs-toggle="tooltip" data-bs-title="number of days in the year"}. 

We can omit the year from a snap if we replace the [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} with a subtrahend and a difference or just a difference. If needed, we can obtain a snap from a difference using the equations below. First, we subtract the difference from the current [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} to obtain the "[eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} difference difference" ([edd]{.tool data-bs-toggle="tooltip" data-bs-title="eda difference difference"}). Then, we use the [edd]{.tool data-bs-toggle="tooltip" data-bs-title="eda difference difference"} to get the "cycle of era" ([coe]{.tool data-bs-toggle="tooltip" data-bs-title="cycle of era"}), "day of cycle" ([doc]{.tool data-bs-toggle="tooltip" data-bs-title="day of cycle"}), "year of cycle" ([yoc]{.tool data-bs-toggle="tooltip" data-bs-title="year of cycle"}), and then finally the year and [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"}.

::: {.equationgroup #equationgroup21}
$$\text{edd} = \text{eda} - \text{difference}$$ {#eq-edd}

$$\text{coe} = \Biggl \lfloor \frac{\begin{cases}\text{edd}&{\text{if } \text{edd} \geq 0;}\\\text{edd}-146096&{\text{otherwise.}}\end{cases}}{146097} \Biggr \rfloor$$ {#eq-coe}

$$\text{doc} = \text{edd} - \text{coe} \times 146097$$ {#eq-doc}

$$\text{yoc} = \biggl \lfloor \frac{\text{doc} - \lfloor \frac{\text{doc}}{1460} \rfloor + \lfloor \frac{\text{doc}}{36524} \rfloor - \lfloor \frac{\text{doc}}{146096} \rfloor}{365} \biggr \rfloor$$ {#eq-yocedd}

$$\text{year} = \text{yoc} + \text{coe} \times 400$$ {#eq-year}

$$\text{ada} = \text{doc} - \text{yoc} \times 365 - \lfloor \frac{\text{yoc}}{4} \rfloor + \lfloor \frac{\text{yoc}}{100} \rfloor$$ {#eq-adadoc}
:::

In the equations above, [eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"} is "epochal day aggregate", [coe]{.tool data-bs-toggle="tooltip" data-bs-title="cycle of era"} is "cycle of era", [doc]{.tool data-bs-toggle="tooltip" data-bs-title="day of cycle"} is "day of cycle", and [yoc]{.tool data-bs-toggle="tooltip" data-bs-title="year of cycle"} is "year of cycle".

is "epochal day aggregate" ([eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"}), If a subtrahend does not include a and a difference We do not need to include a  with , because  can also omit the .
The current Zone [0]{.color0} [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"}+[ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} timestamp is \${zeroYearHsl}+\${zeroAdaHsl1}.


omit the [yoe]{.tool data-bs-toggle="tooltip" data-bs-title="year of era"}  by passing the subtrahend, difference, and current "epochal day aggregate" ([eda]{.tool data-bs-toggle="tooltip" data-bs-title="epochal day aggregate"}) to the 

We can omit the [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} if the timestamp is based on Zone [0]{.color0}. 
 The time since this webpage was loaded is current - load = diff.
If needed, calculate the year by subtracting the [ada]{.tool data-bs-toggle="tooltip" data-bs-title="annual day aggregate"} and &Delta; from the current eda and passing the result to the equations below.


To make it easier to compare timestamps, we should agree to always set the [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} to zero. If we use a non-zero [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"}, we should include it in between the event time and the elapsed time.

Let’s say that you live in Zone [0]{.color0} and there are four [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"} that are vital to your typical daily rhythm: you start work at [375]{.color375} md, take a lunch break from [525]{.color525} to [575]{.color575} md, and finish work at [725]{.color725} md. After one event passes, you can begin tracking the time until the next one and thus cycle through all four over the course of each day: current = [375]{.color375} diff = [525]{.color525} diff = [575]{.color575} diff = [725]{.color725} diff.

Tzo range input (include tzo only if not zero)

Another use case for minuend expansion is travel. If the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is the minuend and your estimated arrival time is the subtrahend, then travel time remaining will be the difference. When traveling, we can replace time with distance so that distance traveled so far is the minuend, the total distance is the subtrahend, and the distance remaining is the difference. Apart from travel, measuring distance in addition to time can be useful for tracking exercise such as running, bicycling, or swimming.

To measure distances, Dec uses taurs ([c]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}) or zems ([z]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}), with or without metric prefixes, depending on the order of magnitude of the distance being measured. One [c]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is close to the circumference of the earth and is equal to 10^5^ kilozem ([kz]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}) or 4 &times 10^5^ kilometers ([km]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}). The distance between the Earth and the Moon ranges from 9.065 to 10.135 c. For larger distances, we can use astronomical units (au), light years (ly), or multiples of [c]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} such as kilotaurs ([kc]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}) and gigataurs ([gc]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}).

au = 3740 c = 3.74 kc
ly ‎ =  236525000 c = 236525 kc = 2.36525 gc

Speed is distance over time. Dec measures speed in omegar ([v]{}).
The speed of the rotation of the Earth is close to one v and is equal to one c per day or one kz per b. The average speed of the orbit of the Moon around the Earth is about 2.2 v and the mean speed of the  Earth orbit around the Sun is approximately 64 v. For higher speeds, we can use percent (%) or permille (‰) light speed (ls) and multiples of v like kv.

%ls ‎ =  10‰ls ‎ =  647552 v = 647.552 kv

For distances that we can measure on Earth, we may want to use submultiples of c such as mc or nc. One mc is 100 kz or 40 km and one nc is 1 decizem (dz) or 4 centimeters (cm). Similarly, we can use  submultiples of v for speeds on Earth. One milliomegar (mv) is one mc per day, one kz per centiday (cd), or one z per b and is very close to one mile per hour. An airplane flying at an average speed of 500 mv will travel the 12500 kz from Cambridge MA to Cambridge UK in a quarter day: 12500 kz / 500 mv = 25 cd.

Around the world in 80 days is a book published by Jules Verne in 1872. The average speed of a trip around the circumference of the earth in 80 days would be about 1 / 80 v or 12.5 mv, which is a typical speed for a bicycle. If we flew in a “[pedal powered airship](https://en.wikipedia.org/wiki/Airship#:~:text=pedal%20powered%20airship)” from Cambridge MA to Cambridge UK at an average speed of 12.5 mv, we could complete the journey in ten days: 12500 kz / 12.5 mv = 1000 cd.

At sunrise, the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} is the amount of time until sunset. At solar noon, the time remaining before sunset will be [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"}. Throughout the day, the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is equal to the sum of sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and the difference between the current and sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="times of day"}: ${selectedSunset.toFixed(3)} -- ${selectedSunrise.toFixed(3)} = \${parseFloat(daytimeDuration.toFixed(3))}.


Like a countdown sequence, we can keep track of the time relative to sunset throughout the day. Instead of the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}, we can display the , the  and the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is equal to the difference between the sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} and the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} is equal to the difference between the sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} and the [da/2]{.tool data-bs-toggle="tooltip" data-bs-title="half day arc"}. The Given a sunset [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} of We can replace the current [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} with an expression that provides the time relative to a given tod.

can keep track of the remaining time until sunset by minuend The [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} of sunrise is found by subtracting half the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} from solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}. The [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"} of sunset is found by adding half the [da]{.tool data-bs-toggle="tooltip" data-bs-title="day arc"} to solar noon [tod]{.tool data-bs-toggle="tooltip" data-bs-title="time of day"}.

# Sunrise and sunset {#sec-sas .hiddenheading}

## Solar hour angle {#sec-sha .unnumbered .hiddenheading}


To find the UTC [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offset"} of a given longitude and latitude, we could use an [application programming interface](https://en.wikipedia.org/wiki/API#:~:text=a%20type%20of%20software%20interface%2C%20offering%20a%20service%20to%20other%20pieces%20of%20software) (API) or a [database](https://observablehq.com/@jcolot/time-zone-service). If we only have longitude, we need to first round degrees to zero or the nearest multiple of fifteen for whole hour [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"}, 7.5 for half hour [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"}, or 3.75 for quarter hour [tzo]{.tool data-bs-toggle="tooltip" data-bs-title="time zone offsets"} and then divide by fifteen to convert degrees to hours.

# Next {#next .unnumbered}

The next article in the [Dec section](/dec) of my [site](/) shows how we can combine [Dec dates](/dec/date) and times into [Dec snaps](/dec/snap)🫰, which are analogous to the [combined date and time representations](https://en.wikipedia.org/wiki/ISO_8601#Combined_date_and_time_representations) in the [ISO 8601](https://en.wikipedia.org/wiki/ISO_8601#:~:text=an%20international%20standard%20covering%20the%20worldwide%20exchange%20and%20communication%20of%20date%20and%20time%2Drelated%20data) international standard for dates and times. The final article in the Dec section demonstrates how Dec dates, times, and snaps🫰can be paired up to express [time intervals](https://en.wikipedia.org/wiki/ISO_8601#Time_intervals:~:text=the%20intervening%20time%20between%20two%20time%20points) called Dec [spans](/dec/span)🌈.

::: {.timenav}
{{< include /asset/_decnav.qmd >}}
:::

Please spread the good word about Dec using the citation information at the bottom of this article. You may also want to cite the Observable notebooks that I adapted into the clock🕓, bar📊chart, map🗺️, and daylight☀️plot visualizations in this article or the 2014 blog post which proposed a system of 20 decimal time zones, each 5 [centidays]{.tool data-bs-toggle="tooltip" data-bs-title="hundredths of a day"} wide, based on the [Greenwich Meridian](https://en.wikipedia.org/wiki/Prime_meridian_(Greenwich)#:~:text=a%20geographical%20reference%20line%20that%20passes%20through%20the%20Royal%20Observatory%2C%20Greenwich%2C%20in%20London%2C%20England):

-   [Pearson, Tom](https://www.2x2.graphics). [[2013]{.underline}]{}[+[124]{}]{.underline}. “Simple D3 clock.” [\${decYear}]{}+[\${decDate}]{}. [https://observablehq.com/\@d3/simple-clock](https://observablehq.com/@d3/simple-clock){.uri}.
-   [Heyen, Frank](https://fheyen.github.io). [[2021]{.underline}]{}[+[246]{}]{.underline}. “BarChart Clock.” [\${decYear}]{}+[\${decDate}]{}. [https://observablehq.com/\@fheyen/barchart-clock](https://observablehq.com/@fheyen/barchart-clock){.uri}.
-   [Johnson, Ian](https://enjalot.github.io). [[2021]{.underline}]{}[+[090]{}]{.underline}. “Draggable World Map Coordinates Input.” [\${decYear}]{}+[\${decDate}]{}. [https://observablehq.com/\@enjalot/draggable-world-map-coordinates-input](https://observablehq.com/@enjalot/draggable-world-map-coordinates-input){.uri}.
-   [Bridges, Dan](https://www.danbridges.org). [[2021]{.underline}]{}[+[311]{}]{.underline}. “Visualizing Seasonal Daylight.” [\${decYear}]{}+[\${decDate}]{}. [https://observablehq.com/\@dbridges/visualizing-seasonal-daylight](https://observablehq.com/@dbridges/visualizing-seasonal-daylight){.uri}.
-   [Clements, John](https://www.brinckerhoff.org). [[2014]{.underline}]{}[+[091]{}]{.underline}, “Decimal Time Zones.” [\${decYear}]{}+[\${decDate}]{}. <https://www.brinckerhoff.org/blog/2014/05/31/decimal-time-zones>.

{{< include /asset/_cite.qmd >}}

{{< include /asset/_glossary.qmd >}}

{{< include /dec/time/_index.qmd >}}
